LUCID: Learning-Enabled Uncertainty-Aware Certification of Stochastic Dynamical Systems

### 2025-07-31 docs/paper/main.tex
feat: update paper

\section{Introduction}
Embodied forms of AI are on the rise, including applications such as autonomous vehicles, robotics, personalized healthcare, and smart infrastructure.
Their core functionality is build around the advances of deep learning, enabling systems to understand and reason in complex and human-like ways to produce a desired behavior.
Whilst the black-box nature of AI components has been a crucial contributor to the widespread success of deep learning, in response to a rapidly evolving legal scrutinization~\cite{veale2021demystifying}, the resulting lack of traceability and certifiability has put a premature halt to its deployment in safety-critical applications.

As deep learning agents are left to choose their actions autonomously in closed-loop interaction with the physical world, they are confronted with a world riddled with randomness and uncertainty. 
Efforts to establishing trust in their safe operation have been largely focused on developing tools to verify the input--output behavior of \glspl{nn} embedded in the systems~\cite{liu2021algorithms}.
Unfortunately, there do not exist any tools that could take these results and certify the safety of the entire system, as uncertainty-aware models of closed-loop systems are rarely available and simulators are often too complicated and opaque to perform verification directly~\cite{wongpiromsarn2023formal}. This motivates a holistic black-box treatment~\cite{corso2021survey}, where safety guarantees are to be established from behavioral data and in account of the unknown uncertainty.

There exist no tools capable of quantifying safety guarantees for complex stochastic closed-loop systems from data.
%
For the setting where a model of the closed-loop system is available, \npinterval by \citet{harapanahalli2023forward} verifies safety of nonlinear systems with non-deterministic disturbance based on existing \gls{nn} verification tools. See the references therein and the annual friendly competition by \citet{abate2024arch} for adjacent work.
%
Few data-driven approaches for stochastic systems exist. 
\omnisafe is a comprehensive platform for the development of safe \gls{rl} algorithms~\citet{ji2024omnisafe}. However, safe \gls{rl} does generally only \emph{encourage} safer behavior, without providing any rigorous or quantified guarantees on the probability of the absence of unsafe behavior.
As a popular working principle for certifying safety, \glspl{cbc}~\cite{prajna2006barrier} are at the basis of tools such as \trust~\cite{gardner2025trust} (supports only polynomial dynamics) and \fossil~\cite{edwards2024fossil} (model-based), which are restricted to deterministic systems.

To close this gap (see Table~\ref{tab:qualitative_comparison}), we introduce \lucid, the first verification engine for black-box stochastic dynamical systems with quantified guarantees.
\begin{table}
    \begin{tabular}{ccccccc}
        \toprule
        \textbf{Tool}                                    & \multicolumn{6}{c}{\textbf{Supported Features}}                                                                                                                                                             \\[.4em]
                                                         & \rotatebox{90}{Guarantees}                      & \rotatebox{90}{Data Driven} & \rotatebox{90}{Stochastic D.} & \rotatebox{90}{Non-Poly. D.} & \rotatebox{90}{Stat. Correct.} & \rotatebox{90}{Closed Loop} \\
        \midrule
        \textbf{\lucid}                                  & \cmark                                          & \cmark                      & \cmark                        & \cmark                       & \cmark                         & \cmark                      \\
        \trust~\shortcite{gardner2025trust}              & \cmark                                          & \cmark                      & \xmarkred                     & \xmarkred                    & \cmark\footnotemark[1]            & \cmark                      \\
        \fossil~\shortcite{edwards2024fossil}            & \cmark                                          & \xmarkred                   & \xmarkred                     & \cmark                       & NA                             & \cmark                      \\
        \omnisafe~\shortcite{ji2024omnisafe}             & \xmarkred                                       & \cmark                      & \cmark                        & \cmark                       & \xmarkred                      & \cmark                      \\
        \npinterval~\shortcite{harapanahalli2023forward} & \cmark                                          & \xmarkred                   & \cmark/\xmarkred\footnotemark[2] & \cmark                            & NA                             & \cmark                      \\
        % \deepsplit~\shortcite{henriksen2021deepsplit} & \xmarkred & \xmarkred & \cmark & \xmarkred& \xmarkred \\
        % \reluplex~\shortcite{katz2017reluplex} & \xmarkred & \xmarkred & \cmark & \xmarkred & \xmarkred \\
        % tool6                                            &                                                 &                             &                               &                                                                                             \\
        \bottomrule
    \end{tabular}
    \caption{Qualitative comparison with existing tools based on their supported features: quantified safety guarantees, data driven, stochastic dynamics, non-polynomial dynamics, statistical correctness guarantees of the learned model (only applicable if data driven), and support for closed-loop systems.}
    \label{tab:qualitative_comparison}
\end{table}
\footnotetext[1]{Assuming the data satisfies persistence of excitation.} % TRUST
\footnotetext[2]{Accepts non-deterministic bounded disturbances.} % npinterval
At its core, it learns \glspl{cbc} for unknown systems based solely on data, by constructing an uncertainty-aware estimator of the expected system behavior based on \glspl{cme}~\cite{muandet2017kernel}.
Crucially, the underlying learning framework establishes quantified safety probabilities with distributionally robust guarantees for arbitrary smooth dynamics, thus relaxing the need for restrictive structural assumptions common to related approaches.
For instance, \citet{schoen2024distributionally} learn \glspl{cbc} for systems with polynomial dynamics and \citet{chen2025distributionally} assume systems with known deterministic component.
%
Alternative approaches often leave the statistical correctness with respect to the underlying data-generating process unaddressed: \citet{kazemi2020fullLTL} use model-free reinforcement learning, relying on known Lipschitz constants; \citet{lew2021sampling} compute reachable sets for stochastic systems using a sampling-based scheme, which yields only asymptotic guarantees; and \citet{salamati2024data} use a scenario-based method, based on Lipschitz constants and exponentially large datasets.
For data-driven safety verification of stochastic systems via conformal prediction~\cite{lindemann2024formal}, no tools are available.

@@

\paragraph{Organization}
The paper is organized as follows:
Section~\ref{sec:theory} provides the theoretical foundations of \lucid, including the safety of black-box dynamical systems, control barrier certificates, conditional mean embeddings, and derivation of the relaxed linear program.
Section~\ref{sec:tool} describes the architecture and functionalities of \lucid alongside a running example, presenting \lucid's components and how they interact.
% Section~\ref{sec:example} presents a simple example use case, demonstrating how to use \lucid to certify the safety of a simple linear system.
Section~\ref{sec:experiments} evaluates \lucid on a suite of benchmarks. Finally, Sections~\ref{sec:future_extensions} and \ref{sec:conclusion} provide future extensions and a conclusion.

@@

\paragraph{Dataset}
Due to their complexity and opacity, such systems must often be treated as black boxes in their entirety, assuming only access to a finite amount, $N$, of system observations
\begin{equation}
    \data_N\colon \quad \{x^i,a^i,x^i_+\}_{i=1}^N,\label{eq:data}
\end{equation}
where every sample transition from $x^i\in\X$ to a successor $x^i_+\in\X$ is generated as a realization
\begin{equation*}
    (x^i,a^i,x_+^i)\sim\int \delta_{f(x^i,a^i,w)}\,p_w(dw)\,\mathcal{U}_\X(dx^i)\,\mathcal{U}_\A(da^i),
\end{equation*}
with $\mathcal{U}_\X,\mathcal{U}_\A$ uniform distributions on $\X$ and $\A$, respectively.

@@

\paragraph{Safety Problem}
A question that often arises when designing a policy $\pi$ for a system $\S$, especially in engineering contexts, is if the closed-loop system $\S^\pi$ elicits safe behavior. That is, when operating in some operating domain $\X_{0}\subset\X$ under a policy $\pi$, the system $\S^\pi$ shall avoid unsafe regions $\X_U\subset\X$ (such as obstacles) for at least some predefined control horizon $T\in\N\cup\{\infty\}$.

Such safety problems are tricky and even for the case where an explicit model \eqref{eq:model} is given, certifying safety of a discrete-time stochastic system over an uncountable state space does not generally admit an analytical solution and is extremely challenging, especially for complex dynamics~\cite{abate2008probabilistic}.
Even more, for unbounded stochasticity $w_t\sim p_w$, there is generally no deterministic answer (safe or not), but a safety probability $P_{\text{safe}}(\S^\pi)\in[0,1)$. Thus the goal is generally to estimate a \emph{lower bound} on the true safety probability $P_{\text{safe}}(\S^\pi)$.

@@

A non-negative function $\B\colon\X\rightarrow\R_{\geq 0}$ is a \gls{cbc} of a system $\S$ with reference to an unsafe set $\X_U$, if it satisfies
\begin{itemize}
    \item[(a)] $\forall x_0\in \X_0\colon\,\B(x_0)\leq\eta$;
    \item[(b)] $\forall x_U\in \X_U\colon\,\B(x_U)\geq\gamma$; and
    \item[(c)] $\forall x\in\X,\,\exists a\in\A\colon\,\E[\B(X_+) \mid X=x,\,A=a] -\B(x) \leq  c;$
\end{itemize}
for some constants $\gamma>\eta\geq0$ and $c\geq 0$. Here, upper case $X_+$, $X$, and $A$ denote the random variables underlying concrete realizations $(x_t,a_t,x_{t+1})$ elicited by $\S$.
%
Intuitively, if one can find a \gls{cbc} for a system $\S$, then, a lower bound on the probability of $\S$ being safe can be quantified based on the distance between the two level sets $\gamma$ and $\eta$~\cite{kushner1967stochastic}:
\begin{equation}
    P_{\text{safe}}(\S^\pi)\geq 1- \frac{\eta + cT}{\gamma},\label{eq:safety_prob_lb}
\end{equation}
where $T$ is the desired control horizon.
Notably, the system dynamics only enters the barrier conditions through the conditional expectation in (c), and a \gls{cbc} thus only depends on the \emph{expected} system behavior, with no need to consider its concrete stochastic law.

@@

\subsection{Data-Driven Dynamics Estimation via Conditional Mean Embeddings}
To reason about the expected value of a random variable, embedding the variable into a (higher dimensional) space and forming a data-driven estimate is a well-established concept in machine learning \cite{scholkopf2002learning,Steinwart2008SVM}.
Following the same reasoning, the \emph{kernel mean embedding} represents the projection of a probability measure into a \gls{rkhs} via a feature map $\phi$ associated with a positive definite kernel $k$~\cite{Smola2007EmbedDistrb,muandet2017kernel}.
For conditional probability distributions a similar concept exists: \glspl{cme} can be used to model the expected value of any \gls{rkhs} function $f\colon\X\rightarrow\R$ under a stochastic process such as $\S$~\cite{park2020measuretheoretic,muandet2017kernel}.
For the Gaussian kernel,
\begin{equation}
    k(x,x') := \sigma_f^2 \exp\left( -\tfrac{1}{2} (x-x')\T \Sigma\, (x-x') \right),\label{eq:gaussian_kernel}
\end{equation}
where $\Sigma:=\diag(\sigma_l)^{-2}$, with hyperparameters $\sigma_f,\sigma_l\in\R$, this encompasses all smooth functions $f$.
A data-driven estimate can then be obtained in closed form from a finite amount of data $\data_N$:
\begin{align}
    \begin{split}
         & \E[f(X_+)\mid X=x,\,A=a] \approx                                                  \\
        % \innerH{f}{\hat{\Psi}^N_{X^+|X}(x)}{\Hilbert_{k}} = 
         & \hspace{5em} k_{XA}^N(x,a)\T\left[ K_{XU}^N+N \lambda I_N\right]^{-1}\! f(X_+^N),
    \end{split}\label{eq:empiricalEstimate}
\end{align}
with column vector $k_{XA}^N(x,a):=[k((x^i,a^i),(x,a))]_{i=1}^N$, Gram matrix $K_{XA}^N:=[k((x^i,a^i),(x^j,a^j))]_{i,j=1}^N$, regularization factor $\lambda\geq0$, identity matrix $I_N$, and $f(X_+^N):=[f(x^i_+)]_{i=1}^N$.
The empirical estimator in \eqref{eq:empiricalEstimate} converges in expectation to the true \gls{cme} for $N\rightarrow\infty$~\cite{park2020measuretheoretic}.

It is common practice to robustify empirical estimates such as \eqref{eq:empiricalEstimate} to out-of-sample behavior by constructing an \gls{rkhs} ambiguity set centered at the empirical \gls{cme}. The result is a distributionally robust estimator with an adjustable robustness radius. The details are omitted here for brevity, but the interested reader is referred to~\citet{kuhn2025distributionally,li2022optimal}.

\subsection{Data-Driven Spectral Barriers}\label{sec:ddbarriers}
Based on the data-driven estimator in \eqref{eq:empiricalEstimate}, the problem of computing \glspl{cbc} based on data can be formulated as a nonconvex semi-infinite program, which for general classes of systems and barriers is extremely difficult to solve.
To arrive at a tractable solution, \lucid conducts two additional steps:
\paragraph{1. Spectral Abstraction} Inspired by the popular random Fourier features approach by \citet{rahimi2007random}, the Gaussian kernel \eqref{eq:gaussian_kernel} admits a Fourier expansion
\begin{align}
    k(x, x') \equiv \sigma_f^2 \int_{\R^{n}}\!\! \mathcal{N}(d\omega\mid0,\Sigma)\; e^{\mathbf{i}\omega\T (P(x)-P(x'))},\label{eq:sqexp_kernel_fourier}
\end{align}
with the zero-mean Gaussian distribution $\smash{\mathcal{N}(d\omega\mid0,\Sigma)}$ with covariance $\Sigma$, the imaginary unit $\mathbf{i}:=\smash{\sqrt{-1}}$, and where the projection $x \mapsto P(x)$ maps the domain $\X$ into the unit hypercube $[0,1]^{n}$.
Partitioning the space of spatial frequencies $\omega\in\R^n$ into a set of discrete frequency bands (see Figure~\ref{fig:spectral_barrier}) yields a spectral abstraction of the associated \gls{rkhs}, i.e., characterizing learnable functions $B$ in the form of Fourier series.
By truncating the series to a finite number, $M$, of fixed frequency bands $\omega_j\in\R^{n}$, $j\in\{0,\ldots,M\}$, the resulting barriers are of the form
\begin{align*}
    \B(x) = \alpha_0 +\sum_{i=1}^{M} \alpha_i \cos\left(\omega_i\T P(x)\right) + \beta_i \sin\left(\omega_i\T P(x)\right).%\label{eq:blackbox:representer_form_spectral}
\end{align*}
Notably, $\B$ admits the linear form $B(x)=b\T\phi_M(x)$, with a truncated Fourier feature map $\phi_M\colon\X\rightarrow\R^{2M+1}$, parametrized by learnable spectral amplitudes
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{fourier_series.pdf}
    \caption{Spectral barrier certificate $B(x)=b\T\phi_M(x)$}
    \label{fig:spectral_barrier}
\end{figure}
\begin{align*}
    b\!=\!\begin{bmatrix}
              \frac{\alpha_0}{\sigma_f^2w_0^2}\! & \frac{\alpha_1}{2\sigma_f^2w_1^2}\!\! & \frac{\beta_1}{2\sigma_f^2w_1^2}\!\! & \ldots\!\! & \frac{\alpha_{M}}{2\sigma_f^2w_{M}^2}\!\! & \frac{\beta_{M}}{2\sigma_f^2w_{M}^2}
          \end{bmatrix}\T\!\!\!\!\in\R^{2M+1}\!\!,
\end{align*}
where the weights $w_0,\ldots,w_{M}\in\R_{\geq 0}$ associated with each frequency band are determined efficiently from the kernel's Gaussian spectral measure via the multivariate CDF (see Figure~\ref{fig:spectral_measure_abstraction}).
The \gls{cme}-based estimator \eqref{eq:empiricalEstimate} is approximated in the same finite basis via $H\in\R^{(2M+1)\times (2M+1)}$ such that
\begin{equation*}
    k_{XA}^N(x,a)\T\left[ K_{XU}^N+N \lambda I_N\right]^{-1}\! \Phi_{M,+}^N
    \approx
    \phi_M{\color{red}(x,a)}\T H,
\end{equation*}
where $\Phi_{M,+}^N:=[\phi_M(x_+^i)\T]_{i=1}^N$,
with an approximation error decreasing exponentially with $M$~\cite{rahimi2007random}.
This reduces the barrier synthesis problem
% This simplifies the barrier problem 
to a semi-infinite linear program, with barrier conditions linear in $b$:
\begin{itemize}
    \item[(a)] $\forall x_0\in \X_0\colon\,\phi_M(x_0)\T b\leq\eta$;
    \item[(b)] $\forall x_U\in \X_U\colon\,\phi_M(x_U)\T b\geq\gamma$; and
    \item[(c)] $\forall x\in\X,\,\exists a\in\A\colon\,\phi_M{\color{red}(x,a)}\T (Hb -b) \leq  c.$
\end{itemize}
% \OS{Importantly, the closed-form estimator in \eqref{eq:empiricalEstimate} does not depend on the choice of features and only through the spectral abstraction is the feature map introduced directly.}
\begin{figure}
    \centering
    % GAUSSIANs: 68-95-99 rule
    \begin{tikzpicture}
        \def\N{50} % number of sample points
        \def\B{0};
        \def\Bs{3.0};
        \def\var{1.3}
        \def\xmax{\B+3.5*\Bs};
        \def\ymin{{-0.1*gauss(\B,\B,\Bs)}};
        \def\h{0.08*gauss(\B,\B,\Bs)};

@@

Since we know the expected behavior of the example system $\S$ from~\eqref{eq:linear-system}, we can run the formal validation by adding the following two lines to the configuration:
\begin{lstlisting}[language=yaml,backgroundcolor=\color{ipython_bg}]
system_dynamics: ["x1 / 2"]
verify: true
\end{lstlisting}
Doing so confirms that the $\B$ is indeed a valid \gls{cbc} for the system.

@@

\Sadegh{Are we comparing with alternative approaches in the literature? Since we mention alternative tools and their capabilities in Table 1, this raises expectations that we provide a fair benchmarking and show how we are better than them.}

>>>>

\section{Introduction}
Embodied forms of AI are on the rise, including applications such as autonomous vehicles, robotics, personalized healthcare, and smart infrastructure.
Their core functionality is built around the advances of deep learning, enabling systems to understand and reason in complex and human-like ways to produce a desired behavior.
Whilst the black-box nature of AI components has been a crucial contributor to the widespread success of deep learning, in response to a rapidly evolving legal scrutinization~\cite{veale2021demystifying}, the resulting lack of traceability and certifiability has put a premature halt to its deployment in safety-critical applications.

As deep learning agents are left to choose their actions autonomously in closed-loop interaction with the physical world, they are confronted with a world riddled with randomness and uncertainty. 
Efforts to establishing trust in their safe operation have been largely focused on developing tools to verify the input--output behavior of \glspl{nn} embedded in the systems~\cite{liu2021algorithms}.
Unfortunately, there do not exist any tools that could take these results and certify the safety of the entire system, as uncertainty-aware models of closed-loop systems are rarely available and simulators are often too complicated and opaque to perform verification directly~\cite{wongpiromsarn2023formal}. This motivates a holistic black-box treatment~\cite{corso2021survey}, where safety guarantees are to be established from behavioral data and in account of the unknown laws of randomness themselves.

There exist no tools capable of quantifying safety guarantees for complex stochastic closed-loop systems from data.
%
For the setting where a model of the closed-loop system is available, \npinterval by \citet{harapanahalli2023forward} verifies safety of nonlinear systems with non-deterministic disturbance based on existing \gls{nn} verification tools. See the references therein and the annual friendly competition by \citet{abate2024arch} for adjacent work.
%
Few data-driven approaches for stochastic systems exist. 
\omnisafe is a comprehensive platform for the development of safe \gls{rl} algorithms~\cite{ji2024omnisafe}. However, safe \gls{rl} does generally only \emph{encourage} safer behavior, without providing any rigorous or quantified guarantees on the probability of the absence of unsafe behavior.
As a popular working principle for certifying safety, \glspl{cbc}~\cite{prajna2006barrier} are at the basis of tools such as \trust~\cite{gardner2025trust}, which supports only polynomial dynamics, and \fossil~\cite{edwards2024fossil}, which is model-based, and both being restricted to deterministic systems.

To close this gap (see Table~\ref{tab:qualitative_comparison}), we introduce \lucid, the first verification engine for black-box stochastic dynamical systems with quantified guarantees.
At its core, \lucid learns \glspl{cbc} for unknown systems based solely on data, by constructing an uncertainty-aware estimator of the expected system behavior based on \glspl{cme}~\cite{muandet2017kernel}.
Crucially, the underlying learning framework establishes quantified safety probabilities with distributionally robust guarantees for arbitrary smooth dynamics, thus relaxing the need for restrictive structural assumptions common to related approaches.
For instance, \citet{schoen2024distributionally} learn \glspl{cbc} for systems with polynomial dynamics and \citet{chen2025distributionally} assume systems with known deterministic component.
\begin{table}
    \centering
    \begin{tabular}{ccccccc}
        \toprule
        \textbf{Tool}                                    & \multicolumn{6}{c}{\textbf{Supported Features}}                                                                                                                                                             \\[.4em]
                                                         & \rotatebox{90}{Guarantees}                      & \rotatebox{90}{Data Driven} & \rotatebox{90}{Stochastic Dyn.} & \rotatebox{90}{Non-Poly. Dyn.} & \rotatebox{90}{Stat. Correct.} & \rotatebox{90}{Closed Loop} \\
        \midrule
        \textbf{\lucid}                                  & \cmark                                          & \cmark                      & \cmark                        & \cmark                       & \cmark                         & \cmark                      \\
        \trust~\shortcite{gardner2025trust}              & \cmark                                          & \cmark                      & \xmarkred                     & \xmarkred                    & \cmark\footnotemark[1]            & \cmark                      \\
        \fossil~\shortcite{edwards2024fossil}            & \cmark                                          & \xmarkred                   & \xmarkred                     & \cmark                       & NA                             & \cmark                      \\
        \omnisafe~\shortcite{ji2024omnisafe}             & \xmarkred                                       & \cmark                      & \cmark                        & \cmark                       & \xmarkred                      & \cmark                      \\
        \npinterval~\shortcite{harapanahalli2023forward} & \cmark                                          & \xmarkred                   & \cmark/\xmarkred\footnotemark[2] & \cmark                            & NA                             & \cmark                      \\
        % \deepsplit~\shortcite{henriksen2021deepsplit} & \xmarkred & \xmarkred & \cmark & \xmarkred& \xmarkred \\
        % \reluplex~\shortcite{katz2017reluplex} & \xmarkred & \xmarkred & \cmark & \xmarkred & \xmarkred \\
        % tool6                                            &                                                 &                             &                               &                                                                                             \\
        \bottomrule
    \end{tabular}
    \caption{Qualitative comparison with existing tools based on their supported features: quantified safety guarantees, data driven, stochastic dynamics, non-polynomial dynamics, statistical correctness guarantees of the learned model (only applicable if data driven), and support for closed-loop systems.}
    \label{tab:qualitative_comparison}
\end{table}
\footnotetext[1]{Assuming the data satisfies persistence of excitation.} % TRUST
\footnotetext[2]{Accepts non-deterministic bounded disturbances.} % npinterval

%
Alternative approaches often leave the statistical correctness with respect to the underlying data-generating process unaddressed: \citet{kazemi2020fullLTL} use model-free reinforcement learning, relying on known Lipschitz constants; \citet{lew2021sampling} compute reachable sets for stochastic systems using a sampling-based scheme, which yields only asymptotic guarantees; and \citet{salamati2024data} use a scenario-based method, based on Lipschitz constants and exponentially large datasets.
For data-driven safety verification of stochastic systems via conformal prediction~\cite{lindemann2024formal}, no tools are available.

@@

\paragraph{Organization}
The paper is organized as follows.
Section~\ref{sec:theory} provides the theoretical foundations of \lucid, including the safety of black-box dynamical systems, control barrier certificates, conditional mean embeddings, and derivation of the relaxed linear program.
Section~\ref{sec:tool} describes the architecture and functionalities of \lucid alongside a running example, presenting \lucid's components and how they interact.
% Section~\ref{sec:example} presents a simple example use case, demonstrating how to use \lucid to certify the safety of a simple linear system.
Section~\ref{sec:experiments} evaluates \lucid on a suite of benchmarks. Finally, Section~\ref{sec:conclusion_and_future_extensions} provides a conclusion and future extensions.

@@

\paragraph{Dataset}
Due to their complexity and opacity, such systems must often be treated as black boxes in their entirety, assuming only access to a finite amount, $N$, of system observations
\begin{equation}
    \data_N\colon \quad \{(x^i,a^i,x^i_+)\}_{i=1}^N,\label{eq:data}
\end{equation}
where every sampled transition from $x^i\in\X$ to a successor $x^i_+\in\X$ is generated as a realization
\begin{equation*}
    (x^i,a^i,x_+^i)\sim\int \delta_{f(x^i,a^i,w)}(dx_+^i)\,p_w(dw)\,\mathcal{U}_\X(dx^i)\,\mathcal{U}_\A(da^i),
\end{equation*}
with $\mathcal{U}_\X$ and $\mathcal{U}_\A$ uniform distributions on $\X$ and $\A$, respectively, and $\delta$ indicating the Dirac delta distribution capturing the system dynamics.

@@

\paragraph{Safety Problem}
A question that often arises when designing a policy $\pi$ for a system $\S$, especially in engineering contexts, is if the closed-loop system $\S^\pi$ elicits safe behavior. That is, when starting from a domain $\X_{0}\subset\X$ under a policy $\pi$, the system $\S^\pi$ shall avoid unsafe regions $\X_U\subset\X$ (such as obstacles) for at least some predefined time horizon $T\in\N\cup\{\infty\}$.

%Such safety problems are tricky and even for the case where an explicit model \eqref{eq:model} is given, 
Certifying safety of a discrete-time stochastic system over an uncountable state space does not generally admit an analytical solution and is extremely challenging, especially for complex dynamics~\cite{abate2008probabilistic}.
Furthermore, for unbounded stochasticity $w_t\sim p_w$, there is generally no yes/no answer to safety, but a safety probability $P_{\text{safe}}(\S^\pi)\in[0,1)$. Thus the goal is generally to estimate a \emph{lower bound} on the true safety probability $P_{\text{safe}}(\S^\pi)$.

@@

A non-negative function $\B\colon\X\rightarrow\R_{\geq 0}$ is a \gls{cbc} of a system $\S$ with reference to an unsafe set $\X_U$ if it satisfies
\begin{itemize}
    \item[(a)] $\forall x_0\in \X_0\colon\,\B(x_0)\leq\eta$;
    \item[(b)] $\forall x_U\in \X_U\colon\,\B(x_U)\geq\gamma$; and
    \item[(c)] $\forall x\in\X,\,\exists a\in\A\colon\,\E[\B(X_+) \mid X=x,\,A=a] -\B(x) \leq  c;$
\end{itemize}
for some constants $\gamma>\eta\geq0$ and $c\geq 0$. Here, upper case $X_+$, $X$, and $A$ denote the random variables underlying concrete realizations $(x_t,a_t,x_{t+1})$ elicited by $\S$.
%
Intuitively, if one can find a \gls{cbc} for a system $\S$, then, a lower bound on the probability of $\S$ being safe can be quantified based on the distance between the two level sets $\gamma$ and $\eta$~\cite{kushner1967stochastic}:
\begin{equation}
    P_{\text{safe}}(\S^\pi)\geq 1- \frac{\eta + cT}{\gamma},\label{eq:safety_prob_lb}
\end{equation}
where $T$ is the desired time horizon.
Notably, the system dynamics only enters the barrier conditions through the conditional expectation in (c), and a \gls{cbc} thus only depends on the \emph{expected} system behavior, with no need to consider its concrete stochastic law.

@@

\subsection{Data-Driven Dynamics Estimation via Conditional Mean Embeddings}
To reason about the expected value of a random variable, embedding the variable into a (higher dimensional) space and forming a data-driven estimate is a well-established concept in machine learning \cite{scholkopf2002learning,Steinwart2008SVM}.
Following the same reasoning, the \emph{kernel mean embedding} represents the projection of a probability measure into a \gls{rkhs} via a feature map $\phi$ associated with a positive definite kernel $k$~\cite{Smola2007EmbedDistrb,muandet2017kernel}.
For conditional probability distributions a similar concept exists: \glspl{cme} can be used to model the expected value of any \gls{rkhs} function $\new{g}\colon\X\rightarrow\R$ under a stochastic process such as $\S$~\cite{park2020measuretheoretic,muandet2017kernel}.
For the Gaussian kernel,
\begin{equation}
    k(x,x') := \sigma_f^2 \exp\left( -\tfrac{1}{2} (x-x')\T \Sigma\, (x-x') \right),\label{eq:gaussian_kernel}
\end{equation}
where $\Sigma:=\diag(\sigma_l)^{-2}$, with hyperparameters $\sigma_f,\sigma_l\in\R$, \new{the associated RKHS encompasses all smooth functions $g$}.
A data-driven estimate can then be obtained in closed form from a finite amount of data $\data_N$:
\begin{align}
    \begin{split}
         & \E[f(X_+)\mid X=x,\,A=a] \approx                                                  \\
        % \innerH{f}{\hat{\Psi}^N_{X^+|X}(x)}{\Hilbert_{k}} = 
         & \hspace{5em} k_{XA}^N(x,a)\T\left[ K_{XA}^N+N \lambda I_N\right]^{-1}\! f(X_+^N),
    \end{split}\label{eq:empiricalEstimate}
\end{align}
with column vector $k_{XA}^N(x,a):=[k((x^i,a^i),(x,a))]_{i=1}^N$, Gram matrix $K_{XA}^N:=[k((x^i,a^i),(x^j,a^j))]_{i,j=1}^N$, regularization factor $\lambda\geq0$, identity matrix $I_N$, and $f(X_+^N):=[f(x^i_+)]_{i=1}^N$.
The empirical estimator in \eqref{eq:empiricalEstimate} converges in expectation to the true \gls{cme} for $N\rightarrow\infty$~\cite{park2020measuretheoretic}.

It is common practice to robustify empirical estimates such as \eqref{eq:empiricalEstimate} to out-of-sample behavior by constructing an \gls{rkhs} ambiguity set centered at the empirical \gls{cme}. The result is a distributionally robust estimator with an adjustable robustness radius. The details are omitted here for brevity, but the interested reader is referred to the works by \citet{kuhn2025distributionally} and \citet{li2022optimal}.

\subsection{Data-Driven Spectral Barriers}\label{sec:ddbarriers}
Based on the data-driven estimator in \eqref{eq:empiricalEstimate}, the problem of computing \glspl{cbc} using data can be formulated as a nonconvex semi-infinite program, which for general classes of systems and barriers is extremely difficult to solve.
To arrive at a tractable solution, \lucid conducts two additional steps:
\paragraph{1. Spectral Abstraction} Inspired by the popular random Fourier features approach by \citet{rahimi2007random}, the Gaussian kernel \eqref{eq:gaussian_kernel} admits a Fourier expansion
\begin{align}
    k(x, x') \equiv \sigma_f^2 \int_{\R^{n}}\!\! \mathcal{N}(d\omega\mid0,\Sigma)\; e^{\mathbf{i}\omega\T (P(x)-P(x'))},\label{eq:sqexp_kernel_fourier}
\end{align}
with the zero-mean Gaussian distribution $\smash{\mathcal{N}(d\omega\mid0,\Sigma)}$ with covariance $\Sigma$, the imaginary unit $\mathbf{i}:=\smash{\sqrt{-1}}$, and where the projection $x \mapsto P(x)$ maps the domain $\X$ into the unit hypercube $[0,1]^{n}$.
Partitioning the space of spatial frequencies $\omega\in\R^n$ into a set of discrete frequency bands (see Figure~\ref{fig:spectral_barrier}) yields a spectral abstraction of the associated \gls{rkhs}, i.e., characterizing learnable functions $B$ in the form of Fourier series.
By truncating the series to a finite number, $M$, of fixed frequency bands $\omega_j\in\R^{n}$, $j\in\{0,\ldots,M\}$, the resulting barriers are of the form
\begin{align*}
    \B(x) = \alpha_0 +\sum_{i=1}^{M} \alpha_i \cos\left(\omega_i\T P(x)\right) + \beta_i \sin\left(\omega_i\T P(x)\right).%\label{eq:blackbox:representer_form_spectral}
\end{align*}
Notably, $\B$ admits the linear form \new{$B(x)=\phi_M(x)\T b$}, with a truncated Fourier feature map $\phi_M\colon\X\rightarrow\R^{2M+1}$, parametrized by learnable spectral amplitudes
\begin{figure}
    \includegraphics[width=\linewidth]{fourier_series.pdf}
    \caption{Spectral barrier certificate \new{$B(x)=\phi_M(x)\T b$}.}
    \label{fig:spectral_barrier}
\end{figure}
\begin{align*}
    b\!=\!\begin{bmatrix}
              \frac{\alpha_0}{\sigma_f^2\w_0^2}\!\! & \frac{\alpha_1}{2\sigma_f^2\w_1^2}\!\!\! & \frac{\beta_1}{2\sigma_f^2\w_1^2}\!\!\! & \ldots\!\!\! & \frac{\alpha_{M}}{2\sigma_f^2\w_{M}^2}\!\!\! & \frac{\beta_{M}}{2\sigma_f^2\w_{M}^2}
          \end{bmatrix}\T\!\!\!\!\in\R^{2M+1}\!\!,
\end{align*}
where the weights $\w_0,\ldots,\w_{M}\in\R_{\geq 0}$ associated with each frequency band are determined efficiently from the kernel's Gaussian spectral measure via the multivariate CDF (see Figure~\ref{fig:spectral_measure_abstraction}).
The \gls{cme}-based estimator \eqref{eq:empiricalEstimate} is approximated in the same finite basis via $H\in\R^{(2M+1)\times (2M+1)}$ such that
\begin{equation*}
    k_{XA}^N(x,a)\T\left[ K_{XA}^N+N \lambda I_N\right]^{-1}\! \Phi_{M,+}^N
    \approx
    \new{\varphi_M(x,a)}\T H,
\end{equation*}
where $\Phi_{M,+}^N:=[\phi_M(x_+^i)\T]_{i=1}^N$, \new{$\varphi_M(x,a)\colon\X\times\A\rightarrow\smash{\R^{2M+1}}$ a feature map augmenting $\phi_M(x)$}, and
with an approximation error decreasing exponentially with $M$~\cite{rahimi2007random}.
This reduces the barrier synthesis problem
% This simplifies the barrier problem 
to a semi-infinite linear program, with barrier conditions linear in $b$:
\begin{itemize}
    \item[(a)] $\forall x_0\in \X_0\colon\,\phi_M(x_0)\T b\leq\eta$;
    \item[(b)] $\forall x_U\in \X_U\colon\,\phi_M(x_U)\T b\geq\gamma$; and
    \item[(c)] $\forall x\in\X,\,\exists a\in\A\colon\,\new{\varphi_M(x,a)}\T (Hb -b) \leq  c.$
\end{itemize}
\new{For the case where $a$ is provided by a given fixed policy $\pi$, $\varphi_M(x,a)$ reduces to $\phi_M(x)$.}
% \OS{Importantly, the closed-form estimator in \eqref{eq:empiricalEstimate} does not depend on the choice of features and only through the spectral abstraction is the feature map introduced directly.}
\begin{figure}
    % GAUSSIANs: 68-95-99 rule
    \begin{tikzpicture}
        \def\N{50} % number of sample points
        \def\B{0};
        \def\Bs{3.0};
        \def\var{1.3}
        \def\xmax{\B+3.5*\Bs};
        \def\ymin{{-0.1*gauss(\B,\B,\Bs)}};
        \def\h{0.08*gauss(\B,\B,\Bs)};

@@

Since we know the expected behavior of the example system $\S$ from~\eqref{eq:linear-system}, we can run the formal validation by adding the following two lines to the configuration:
\begin{lstlisting}[language=yaml,backgroundcolor=\color{ipython_bg}]
system_dynamics: ["x1 / 2"]
verify: true
\end{lstlisting}
Doing so confirms that $\B$ is indeed a valid \gls{cbc} for the system.

@@

%\Sadegh{Are we comparing with alternative approaches in the literature? Since we mention alternative tools and their capabilities in Table 1, this raises expectations that we provide a fair benchmarking and show how we are better than them.}

\section{Conclusion and Future Extensions}\label{sec:conclusion_and_future_extensions}
\Sadegh{we should combine Conclusion and Future extensions into one section and keep it in the main body. We should also mention sparse CME as a way of improving scalability...}
